R for Clinical Study Reports and Submission

### 2021-09-18 project-folder.Rmd
clarify consistency

In this part, we would like to put you as an A&R leader for a clinical project.

@@

A clearly defined project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically a well defined project folder can achieve:

- Consistency
- Reproducibility
- Automation
- Compliance

@@

For R user, you already benefit from a well defined and consistent folder structure. 
That is the R package folder structure. 
Every R package developer is required to follow the same convention to organize their R
functions, before the function package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and user those packages after installing from CRAN.

We recommend to also use R package folder structure to organize analysis scripts 
for clincial trial development. 
Using the R package folder structure to streamline data analysis work has been proposed 
(@marwick2018packaging, @wuanalysis).

In this book, we refer an R package as **project specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

@@

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `R/`: Project specific R functions.
- `man/`: Manual of project specific R functions.

We demonstrate the the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.

A general discussion of the R package folder structure can be found in Chapter 4 of “R packages:
organize, test, document, and share your code” book (@wickham2015r).

@@

>>>>

To deliver A&R results of a clincial tiral project, it is a team work that typically require
collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

@@

A clearly defined clinical project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically a well defined project folder can achieve:

- Consistency: every team work on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
- Automation: automatically check integration of a project.  
- Compliance: reduce compliance issue

We illustrate the R package project folder for A&R project using a demo project called 
[`esubdemo`](https://github.com/elong0527/esubdemo.git). You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) 
the project using RStudio.

@@

For R user, you already benefit from a well defined and consistent folder structure. 
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
Every R package developer is required to follow the same convention to organize their R
functions, before the R package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
There are many good resources to guide developers on R package development.
For example, [Hadley Wickmans' R packages book](https://r-pkgs.org/)

We recommend to use R package folder structure to organize analysis scripts 
for clinical trial development. 
Using the R package folder structure to streamline data analysis work has also been proposed before 
(@marwick2018packaging, @wuanalysis).

In this book, we refer an R package as **project specific R package**. 
The purpose of project specific R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

@@

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `R/`: Project specific R functions.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `man/`: Manual of project specific R functions.

To illustrate the idea, you can open `esubdemo` project by click the `esubdemo.Rproj` project.

For analysis code to create TLFs, we saved them in the `vignettes/` folder. 
In `esubdemo`, we put analysis code we learned from previous chapters into the `vignettes` folder.

> Here the `adam\` folder contain ADaM datasets. 
> The `tlf` folder contain output TLFs in `rtf` and `pdf` format.
> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
> In a real A&R report, you may have different location to save your input and output.

```
vignettes
├── adam
├── tlf
├── tlf-01-disposition.Rmd
├── tlf-02-population.Rmd
├── tlf-03-baseline.Rmd
├── tlf-04-efficacy.Rmd
├── tlf-05-ae-summary.Rmd
└── tlf-06-ae-spec.Rmd
```

In previous chapters, we also create a few helper functions that can be used for different analysis. 
(e.g. `fmt_num` and `count_by`). 
We put those helper functions into the `R/` folder.

> Here `zzz.R` is a template file that ensure startup file `inst/startup.R` is executed. 
> We will revisit startup file while we discuss reproducibility.

```
R
├── count_by.R
├── fmt.R
└── zzz.R
```

The `man` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentations at the header of each function in the `R\` folder. 
Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).

@@

By using the R package folder structure, you can also benefit from many outstanding tools to 
simplify and streamline your workflow.
